---
title: 'Global Trends In Urban Energy Use: The Tropical Shift'
author: "Elliot Cohen, Vijay Modi, Henri Torbey, Yu Tian and Michael Piccirilli"
date: "October 9 2014"
output:
  pdf_document: default
  html_document:
    theme: spacelab
---
<link href="http://kevinburke.bitbucket.org/markdowncss/markdown.css" rel="stylesheet"></link>

*************
```{r global_options, include=FALSE}
library(knitr)
opts_chunk$set(fig.path='Figs/', echo=FALSE, fig.width=8, fig.height=5, cache=TRUE)
options(width=85)
```

```{r initialize, , include=TRUE, message=FALSE}
filepath<-"~/github/Rcommands/"

## read-in csv files from NRLDC html data scapring (Chris Tan and Myf March 2014)
options(stringsAsFactors=FALSE)

## Load libraries
library(knitr)    # create RMD documents
library(rmarkdown)
library(plyr)     # ddply...
library(reshape2) # melt...
library(ggplot2)  # plots
library(scales)   # used in ggplot()
library(xlsx)     # read excel files
library(sm)       # density plots

load("growth.rsav")         # 5-yr average population growth in cities w. pop > 750k (as reported)
load("growth.melt.rsav")    # "" melted
load("pop.rsav")            # 5-yr average population in cities w. pop > 750k (as reported)
load("pop.melt.rsav")       # "" melted
load("growth.decadal.rsav")  # 10-yr population growth (computed for '50-60, ..., '10-20)
load("growth.decadal2.rsav") # 10-yr population growth (computed for '55-65, ..., '15-25)
load("cities.rsav") # synthesized population and growth rate data for global cities with pop > 750k
load("epochal.rsav") # "cities" summarized by 25-yr epoch
```

```{r functions}
## function to create Figure/Table numbers
incCount <- function(inObj, useName) {
    nObj <- length(inObj)
    useNum <- max(inObj) + 1
    inObj <- c(inObj, useNum)
    names(inObj)[nObj + 1] <- useName
    inObj
}
figCount <- c(`_` = 0)
tableCount <- c(`_` = 0)
## Use like this:
# figCounts <- incCount(figCount, "f.randomFigure")

# function to check for NAs and complete cases
check<-function(df){
  NAs<-sum(is.na(df))
  print(paste("NAs:", NAs)) # count NA's
  if (NAs>0) df1[which(is.na(df)), ] # Show NA's, if any.
  cc<-complete.cases(df)  # check for missing values...
  print(paste("Complete Cases:", all(cc)))  # Given a set of logical vectors, are all of the values true?
  }
```

Motivation
-----------
Many of the world's **largest** and **fastest-growing** cities--from Karachi (pop. 14 million; 34.6% increase from 2000-2010) to Delhi (22m; 39.4%), Dhaka (15m, 45.2%), Jakarta (10m; 14.8%), Bangkok (8m, 29.1%), Lagos (11m; 48.2%) and Kinshasa (9m, 55.4%)--are located in South Asia and Sub-Saharan Africa with tropical to sub-tropical climates unlike those of most OECD member cities in the global North. As the tropics/sub-tropics become increasingly urban, inustrial and affluent, it is important to consider how energy demand--particularly for thermal comfort--will evolve differently in these places than it has historically across the OECD. 

To illustrate the potential for vast differences in energy demand for thermal comfort between cities in the global North and cities in the Tropics/Sub-Tropics, consider Delhi, India.  Delhi--with its massive population and very hot climate--is an outlier compared to OECD member cities but typical of South Asia: Peak summer temperatures routinely exceed 40 deg C. (104 F.), and intense heatwaves can approach 50 deg C. (122 F.). Given the huge temperature differential between outdoor (say 104 F.) and desired indoor air temperature (say 72 F), and the thermodynamic fact that energy for cooling scales linearly with the temperature differential, cooling a room in Delhi requires twice as much energy as cooling a room in New York where the summer outdoor-indoor temperature differential is typically half that. 

In addition to higher temperatures, the summer season is also much longer: in the past year, Delhi had over **six** times as many cooling-degree days as New York City (again assuming a desired indoor air temperature of 72 deg F). Compounded by leaky building envelopes in developing world cities (designed for natural ventilation, not air conditiong), intense heat-island effects (typically less green space), and massive population growth, peak electricity demand in cities thourought the developing world could one day surpass that of their neighbors to the north--not only in aggregate terms because of their population, but also *per-capita* due to climate, building design and thermodynamics. 

Implications of Global Energy Service Provision Parity
--------------
If we consider the trajectory of developing world cities as reaching eventual parity with OECD cities, and if we think in terms of energy *service provision* rather than just BTU or KWh, then we begin to appreciate the magnitude of future energy demand (and associated resource consumption and environmental impact). This has important implications not only for regional grid planning and supply reliability, but also the global transition to renewable energy given the limitations of meeting such large and 'peaky' demand with non-dispatchable resources such as wind and solar.

Figure 1 provides a map of urbanization rates for cities worldwide with a population greater than 750,000 (UN 2011). Urbanization rates are clearly highest in South Asia and Sub-Saharan Africa. Figure 2 shows the *distribution* of population growth rates for the same set of cities, grouped by latitude (global North, Tropics and Subtropics). The expected value (e.g. central tendancy) of urbanization rates in the Tropics and Sub-Tropics is clearly distinct from that of the North.

```{r urbanization_data, cache=FALSE, results='hide', eval=FALSE}
# ## Import data
# ## United Nations, Department of Economic and Social Affairs, Population Division
# ## World Urbanization Prospects: The 2011 Revision
# 
# # Average Annual Rate of Change of Urban Agglomerations with 750,000 Inhabitants or More in 2011, by Country, 1950-2025 (per cent)
# growth=read.xlsx(file="UN_2011_Growth_Rate_Cities_Over_750k.xlsx",
#                  sheetName="GROWTH-RATE-CITIES",
#                  as.data.frame=TRUE,header=TRUE,check.names=TRUE,
#                  startRow=13)
# 
# # Population of urban agglomerations with 750,00 inhabitants or more, 1950-2025
# pop=read.xlsx(file="UN_2011_Population_Cities_Over_750k.xlsx",
#               sheetName="CITIES-OVER-750K",
#               as.data.frame=TRUE,header=TRUE,check.names=TRUE,
#               startRow=13, endRow=646, colIndex=c(1:23))
# 
# # match city/country to continent
# countries<-scan(file="countrylist.txt",
#                 what="charachter", sep="\n",
#                           blank.lines.skip=TRUE)
# 
# dummy<-strsplit(countries, split=" ")
# continent<-laply(dummy, '[[', 1)
# country.code<-laply(dummy, '[[', 4)
# country.name<-laply(dummy, '[[', 5)
# df<-data.frame(Country.Code=country.code, Country.Name=country.name, Continent.Abr=continent)
# df$Country.Code<-as.numeric(df$Country.Code)
# 
# df$Continent<-NA
# df$Continent[which(df$Continent.Abr=="AF")]<-"Africa"
# df$Continent[which(df$Continent.Abr=="AS")]<-"Asia"
# df$Continent[which(df$Continent.Abr=="EU")]<-"Europe"
# df$Continent[which(df$Continent.Abr=="NA")]<-"North.America"
# df$Continent[which(df$Continent.Abr=="SA")]<-"South.America"
# df$Continent[which(df$Continent.Abr=="OC")]<-"Oceania"
# df$Continent[which(df$Continent.Abr=="AN")]<-"Antarctica"
# 
# # merge operation
# merge.by<-"Country.Code"
# print(paste("Both data.frames share a unqiue column name:", merge.by %in% names(pop) & merge.by %in% names(df), sep=" "))
# test<-merge(pop, df, by=merge.by)
# test<-subset(test, select=-Continent.Abr)
# test<-subset(test, select=-Country.Name)
# test<-subset(test, select=-Note)
# test<-subset(test, select=c(23, 1:22))
# pop<-test
# remove(test)
# save(pop, file="pop.rsav")
# 
# merge.by<-"Country.Code"
# print(paste("Both data.frames share a unqiue column name:", merge.by %in% names(growth) & merge.by %in% names(df), sep=" "))
# test<-merge(growth, df, by=merge.by)
# test<-subset(test, select=-Continent.Abr)
# test<-subset(test, select=-Country.Name)
# test<-subset(test, select=-Note)
# test<-subset(test, select=c(22, 1:21))
# growth<-test
# remove(test)
# save(growth, file="growth.rsav")
# 
# ## get data for a couple of well known megacities...
# get<-c("Karachi", "Delhi", "Dhaka", "Jakarta", "Krung Thep (Bangkok)", "Lagos", "Kinshasa")
# 
# ## compute decadal growth rates
# # even decades
# test<-pop[, 1:7]
# test$g50.60<-(pop$X1960-pop$X1950)/pop$X1950*100
# test$g60.70<-(pop$X1970-pop$X1960)/pop$X1960*100
# test$g70.80<-(pop$X1980-pop$X1970)/pop$X1970*100
# test$g80.90<-(pop$X1990-pop$X1980)/pop$X1980*100
# test$g90.00<-(pop$X2000-pop$X1990)/pop$X1990*100
# test$g00.10<-(pop$X2010-pop$X2000)/pop$X2000*100
# test$g10.20<-(pop$X2020-pop$X2010)/pop$X2010*100
# test[test$Urban.Agglomeration %in% get, ]
# growth.decadal<-test
# remove(test)
# save(growth.decadal, file="growth.decadal.rsav")
# 
# # odd decades
# test<-pop[, 1:7]
# test$g55.65<-(pop$X1965-pop$X1955)/pop$X1955*100
# test$g65.75<-(pop$X1975-pop$X1965)/pop$X1965*100
# test$g75.85<-(pop$X1985-pop$X1975)/pop$X1975*100
# test$g85.95<-(pop$X1995-pop$X1985)/pop$X1985*100
# test$g95.05<-(pop$X2005-pop$X1995)/pop$X1995*100
# test$g05.15<-(pop$X2015-pop$X2005)/pop$X2005*100
# test$g15.25<-(pop$X2025-pop$X2015)/pop$X2015*100
# test[test$Urban.Agglomeration %in% get, ]
# growth.decadal2<-test
# remove(test)
# save(growth.decadal2, file="growth.decadal2.rsav")
#
## START HERE if loading previously saved dataframes....
# load("pop.rsav")
# load("growth.rsav")
#
# ## melt for ggplot()
# df1<-melt(pop, id.vars=c(1:7))
# sum(is.na(df1))
# 
# ## assign pretty column names
# get<-levels(df1$variable)
# dummy<-strsplit(get, split="X")
# test<-laply(dummy, '[[', 2)
# df1$variable<-factor(df1$variable, labels=test)
# 
# ## check for NA's
# sum(is.na(df1))          # zero NAs
# df1[which(is.na(df1)), ] # zero NAs
# cc<-complete.cases(df1)  # check for missing values...
# sum(cc)==dim(df1)[1]     # zero missing values.
# 
# ## bin growth rates by quantile
# df1$bin<-cut(df1$value, breaks=quantile(df1$value), include.lowest=TRUE)
# df1$bin.name<-factor(df1$bin, labels=c("1st quartile","2nd quartile","3rd quartile","4th quartile")) # assign pretty names
# names(df1)[8:11]<-c("Year","Pop.","Pop.Bin","Pop.Quartile")
# pop.melt<-df1
# 
# ## check for NAs and complete cases
# NAs<-sum(is.na(df1)); NAs           # count NA's
# if (NAs>0) df1[which(is.na(df1)), ] # Show NA's, if any.
# cc<-complete.cases(df1) # check for missing values...
# all(cc)                 # Given a set of logical vectors, are all of the values true?
# 
# ## repeat for growth data
# df2<-melt(growth, id.vars=c(1:7))
# get<-levels(df2$variable)
# dummy<-strsplit(get, split="X")
# period<-laply(dummy, '[[', 2)
# period<-gsub("[.]", "-", period)
# 
# ## bin growth rates by quantile
# df2$variable<-factor(df2$variable, labels=period)
# df2$bin<-cut(df2$value, breaks=quantile(df2$value), include.lowest=TRUE)
# df2$bin.name<-factor(df2$bin, labels=c("1st quartile","2nd quartile","3rd quartile","4th quartile"))
# names(df2)[8:11]<-c("Period","Growth","Growth.Bin","Growth.Quartile")
# growth.melt<-df2
# 
# ## save
# save(pop.melt, file="pop.melt.rsav")
# save(growth.melt, file="growth.melt.rsav")
# 
# # conform population data and growth rate data for a given year
# dummy<-strsplit(as.character(df2$Period), split="-")  # year ending XXXX
# df2$Year<-laply(dummy, "[[", 2)
# str(df2)
# 
# # check for NAs and complete cases
# NAs<-sum(is.na(df2)); NAs           # count NA's
# if (NAs>0) df1[which(is.na(df2)), ] # Show NA's, if any.
# cc<-complete.cases(df2) # check for missing values...
# all(cc)                 # Given a set of logical vectors, are all of the values true?
# 
# # MERGE
# merge.by<-c("Continent", "Country.Code", "Country", "City.Code", "Urban.Agglomeration", "Latitude", "Longitude", "Year")
# print(paste("Both data.frames share a unqiue column name:", merge.by %in% names(df1) & merge.by %in% names(df2), sep=" "))
# 
# cities<-merge(df1, df2, by=merge.by)
# 
# # apply the function
# check(cities)
#
# ## group by geography... (or try clustering by lat-long attributes)
# tropics<-subset(cities, Latitude>=-23.45 & Latitude <=23.45)
# subtropics<-subset(cities, Latitude>=-38 & Latitude <=-23.45 | Latitude>=23.45 & Latitude <=38)
# north<-subset(cities, Latitude>38)
# 
# ## add region attribute to cities
# cities$region<-NA
# t<-which(cities$Latitude >= -23.45 & cities$Latitude <= 23.45)
# cities$region[t]<-"Tropics"
# 
# st<-which(cities$Latitude>= -38 & cities$Latitude < -23.45 | cities$Latitude > 23.45 & cities$Latitude <= 38)
# cities$region[st]<-"Subtropics"
# 
# n<-which(cities$Latitude > 38)
# cities$region[n]<-"North"
# cities$region<-as.factor(cities$region)
# 
# # ## group countries by UN economic classifications
# # D1<-c("Andorra", "Australia", "Bermuda", "Canada", "Iceland", "Israel", "Liechtenstein", "Monaco", "New Zealand", "Norway", "Korea (South)", "San Marino", "Singapore", "Switzerland", "Taiwan")
# # 
# # D2:  "Albania", "Algeria", "Anguilla", "Antigua and Barbuda", "Argentina", "Aruba", "Bahamas", "Bahrain", "Barbados", "Belize", "Bolivia", "Bosnia and Herzegovina", "Botswana", "Brazil", "British Virgin Islands", "Brunei Darussalam", "Bulgaria", "Cameroon", "Cayman Islands", "Chile", "Colombia", "Cook Islands", "Costa Rica", "Cote d'Ivoire", "Croatia", "Cuba"," Czechoslovakia (Former)", "Korea (North)", "Dominica", "Dominican Republic", "Ecuador", "Egypt", "El Salvador", "Fiji", "French Polynesia", "Gabon", "Ghana", "Grenada", "Guatemala", "Haiti", "Honduras", "Indonesia", "Iran", "Iraq", "Jamaica", "Jordan", "Kenya", "Kuwait", "Lebanon", "Libyan Arab Jamahiriya", "Malaysia", Marshall Islands, Mauritius, Mexico, Micronesia, Mongolia, Montserrat , Morocco, Namibia, Nauru, Netherlands Antilles, New Caledonia, Nicaragua, Nigeria, Occupied Palestine, Oman, Pakistan, Palau, Panama, Papua New Guinea, Paraguay, Peru, Philippines, Puerto Rico, Qatar, Romania, Saint Kitts-Nevis, Saint Lucia, Saint Vincent and the Grenadines, Saudi Arabia, Serbia and Montenegro, Seychelles, Somalia, South Africa, Sri Lanka, Sudan, Suriname, Swaziland, Syria, Macedonia, Thailand, Tonga, Trinidad and Tobago, Tunisia, Turkey, Turks and Caicos Islands, United Arab Emirates, Uruguay, Venezuela, Vietnam, Zanzibar, Zimbabwe
# # 
# # D3:	Afghanistan, Angola, Bangladesh, Benin, Bhutan, Burkina Faso, Burundi, Cambodia, Cape Verde, Central African Republic, Chad, Comoros, Congo (Brazzaville), Congo (Kinshasa), Djibouti, Equatorial Guinea, Eritrea, Ethiopia, Gambia, Guinea, Guinea-Bissau, Guyana, Kiribati, Laos, Lesotho, Liberia, Madagascar, Malawi, Maldives, Mali, Mauritania, Mozambique, Myanmar, Nepal, Niger, Rwanda, Samoa, Sao Tome and Principe, Senegal, Sierra Leone, Solomon Islands, Timor-Leste, Togo, Tuvalu, Uganda, Tanzania, Vanuatu, Yemen, Zambia
# # 
# # EU:	Austria, Belgium, Cyprus, Czech Republic, Denmark, Finland, France, Germany, Greece, Hungary, Ireland, Italy, Luxembourg, Malta, Netherlands, Poland, Portugal, Slovakia, Slovenia, Spain, Sweden, United Kingdom
# # 
# # FSU:	Armenia, Azerbaijan, Belarus, Estonia, Georgia, Kazakhstan, Kyrgyzstan, Latvia, Lithuania, Moldova (Republic of), Russian Federation, Tajikistan, Turkmenistan, Ukraine, USSR (Former), Uzbekistan
# #
# ## summarize population growth by decade
# d50<-which(cities$Period %in% c("1950-1955", "1955-1960"))
# d60<-which(cities$Period %in% c("1960-1965", "1965-1970"))
# d70<-which(cities$Period %in% c("1970-1975", "1975-1980"))
# d80<-which(cities$Period %in% c("1980-1985", "1985-1990"))
# d90<-which(cities$Period %in% c("1990-1995", "1995-2000"))
# d00<-which(cities$Period %in% c("2000-2005", "2005-2010"))
# d10<-which(cities$Period %in% c("2010-2015", "2015-2020"))
# d20<-which(cities$Period %in% c("2020-2025"))
# cities$decade<-NA
# cities$decade[d50]<-"Fifties"
# cities$decade[d60]<-"Sixties"
# cities$decade[d70]<-"Seventies"
# cities$decade[d80]<-"Eighties"
# cities$decade[d90]<-"Nineties"
# cities$decade[d00]<-"Oughts"
# cities$decade[d10]<-"Twenty-Teens"
# cities$decade[d20]<-"Twenty-Twenties"
# 
# # summarize population growth by epoch
# e1<-which(cities$Period %in% c("1950-1955", "1955-1960", "1960-1965", "1965-1970", "1970-1975"))
# e2<-which(cities$Period %in% c("1975-1980", "1980-1985", "1985-1990", "1990-1995", "1995-2000"))
# e3<-which(cities$Period %in% c("2000-2005", "2005-2010", "2010-2015", "2015-2020", "2020-2025"))
# cities$epoch<-NA
# cities$epoch[e1]<-"1950-1975"
# cities$epoch[e2]<-"1975-2000"
# cities$epoch[e3]<-"2000-2025"
# 
# # save final data.frame
# save(cities, file="cities.rsav")
#
# epochal<-ddply(cities, .(Country.Code, Country, City.Code, Urban.Agglomeration, Latitude, Longitude, epoch), summarize, End.Pop.=Pop.[5], Annual.Growth=(Pop.[5]-Pop.[1])/Pop.[1]*100/25, Period.Growth= (Pop.[5]-Pop.[1])/Pop.[1]*100)
# check(epochal) # check for NAs and complete cases
# epochal<-na.omit(epochal)
# epochal$epoch<-as.factor(epochal$epoch)
# epochal$Period.Growth.Bin<-cut(epochal$Period.Growth, breaks=quantile(epochal$Period.Growth), include.lowest=TRUE)
# epochal$Period.Growth.Quartile<-factor(epochal$Period.Growth.Bin, labels=c("1st quartile","2nd quartile","3rd quartile","4th quartile"))
# check(epochal) # check for NAs and complete cases
# save(epochal, file="epochal.rsav")
```

```{r urbanization_map, fig.width=12, fig.height=6, echo=FALSE}
# base map
world <- map_data("world")
worldmap <- ggplot(world, aes(x=long, y=lat, group=group)) +
  geom_path()

# CURRENT Population and Urbanization Rates
ggplot() + 
  geom_path(data=world, alpha=0.5, size=0.2, aes(x=long, y=lat, group=group)) +
#   geom_ribbon(data=world, x=world$long, ymin=-23.45, ymax= 23.45, alpha=0.3, fill="green") +
#   geom_ribbon(data=world, x=world$long, ymin=-38, ymax=-23.45, alpha=0.3, fill="blue") +
#   geom_ribbon(data=world, x=world$long, ymin= 23.45, ymax= 38, alpha=0.3, fill="blue") +
  geom_hline(y=-23.45, linetype=2, size=0.2) +
  geom_hline(y= 23.45, linetype=2, size=0.2) +
  geom_hline(y=-38, linetype=3, size=0.2) +
  geom_hline(y=38, linetype=3, size=0.2) +
  geom_point(data=subset(cities, Period=="2010-2015"), aes(x=Longitude, y=Latitude, colour=Growth.Quartile, size=Pop./10^3)) +
#   coord_equal() +
  scale_y_continuous(breaks=(-2:2) * 30) +
  scale_x_continuous(breaks=(-4:4) * 45) +
  labs(colour = 'Urbanization Rate', size="Population (Millions)", title="Rapid Urbanization Throughout the Tropics and Sub-Tropics") +
  theme_bw() +
  scale_color_brewer(palette="Reds") +
  scale_size(range = c(1, 10))
  
# Evolution over time (10-year)
decadal<-ddply(cities, .(Continent, Country.Code, Country, City.Code, Urban.Agglomeration, Latitude, Longitude, decade), numcolwise(mean))

decadal$decade<-factor(decadal$decade, levels=c("Fifties", "Sixties", "Seventies", "Eighties", "Nineties", "Oughts", "Twenty-Teens", "Twenty-Twenties"))

mapWorld <- borders("world", colour="gray40", fill="gray40")

ggplot() + 
  geom_path(data=world, aes(x=long, y=lat, group=group)) +
  mapWorld +
  geom_point(data=decadal, aes(x=Longitude, y=Latitude, colour=Growth, size=Pop./10^3)) + 
  facet_wrap(~decade, nrow=length(levels(factor(decadal$decade)))/2) +
  coord_equal() +
  scale_y_continuous(breaks=(-2:2) * 30) +
  scale_x_continuous(breaks=(-4:4) * 45) +
  labs(colour = 'Annual Population Growth (%)', size="Population (Millions)") +
  theme_bw() +
  scale_size(range = quantile(decadal$Pop., probs=c(0.0, 1)/500)) +
  scale_colour_gradient(low="white", high="red", limits=quantile(decadal$Growth, probs=c(0.0, 1)))
#   scale_size(range = quantile(decadal$Pop., probs=c(0.05, 0.95))/500) +
#   scale_colour_gradient(low="white", high="red", limits=quantile(decadal$Growth, probs=c(0.05, 1)))

# Evolution over time (25 year)
ggplot() + 
  geom_path(data=world, aes(x=long, y=lat, group=group)) +
  geom_point(data=epochal, aes(x=Longitude, y=Latitude, colour=Period.Growth.Quartile, size=End.Pop./10^3)) + 
  facet_wrap(~epoch, nrow=length(levels(epochal$epoch))) +
  facet_wrap(~epoch) + 
  coord_equal() +
  scale_y_continuous(breaks=(-2:2) * 30) +
  scale_x_continuous(breaks=(-4:4) * 45) +
  labs(colour = 'Urbanization Rate', size="Population (Millions)") +
  theme_bw() +
  scale_color_brewer(palette="Reds") +
  scale_size(range = c(2, 10))

## Summary table of top 20 largest cities by population
test<-subset(cities, Period=="2010-2015")
test<-test[order(-test$Pop.), ]
test$Current.Pop.<-test$Pop./10^3

# give shorter names for print layout
usa<-which(test$Country=="United States of America")
test$Country[usa]<-"U.S.A"
uae<-which(test$Country=="United Arab Emirates")
test$Country[uae]<-"U.A.E"
DRC<-which(test$Country=="Democratic Republic of the Congo")
test$Country[DRC]<-"D.R.C"

megacities<-subset(test, Current.Pop.>=10, select=c("Continent","Country","Urban.Agglomeration","Current.Pop."))
rownames(megacities)<-NULL
colnames(megacities)[4]<-"Population [MM]"
print("World's Largest Cities in 2015 (Population > 10 million)")
print(megacities, digits=3)

fastcities<-test[order(-test$Growth), ]
fastcities<-subset(fastcities, select=c("Continent","Country","Urban.Agglomeration","Growth"))
rownames(fastcities)<-NULL
colnames(fastcities)[4]<-"Growth.Rate [%]"
print("World's Fastest Growing Cities (2010-2015) with a Population > 750,000")
print(fastcities[1:20,], digits=3)

remove(test)
```

```{r urbanization_PDF, echo=FALSE, fig.width=7, fig.height=4}
## density plots
test<-subset(cities, Year==2015)
sm.density.compare(test$Growth, test$region, xlab="% Change (Annual)", xlim=quantile(test$Growth, probs=c(0.0,.99)))
title(main="Population Growth Rate of Cities, by Region")
colfill<-c(2:(2+length(levels(test$region)))) 
legend("topright", levels(test$region), fill=colfill)

## also show cooling degree days for cities in these regions..
```

```{r urban_temps, eval=FALSE, cache=TRUE, fig.width=7, fig.height=4, fig.align='left'}
## Load list of cities from Hadley Urban Analysis
# file<-"http://www.metoffice.gov.uk/hadobs/urban/data/Station_list1.txt"
# stns<-read.fwf(file, widths=c(5,18,7,7), header = FALSE, sep = "\t", skip = 5, strip.white=TRUE)
# names(stns)<-c("WMONo", "Stn.name","Lat","Long")
# 
# ## load Hadley ISH station metadata
# load("/Users/elliotcohen/Dropbox/data/Climate/Hadley/metdata.rsav")
# metdata$USAFID<-as.factor(metdata$USAFID)
# 
# ## add region attribute to cities
# metdata$region<-NA
# t<-which(metdata$Lat >= -23.45 & metdata$Lat <= 23.45)
# metdata$region[t]<-"Tropics"
# 
# st<-which(metdata$Lat>= -38 & metdata$Lat < -23.45 | metdata$Lat > 23.45 & metdata$Lat <= 38)
# metdata$region[st]<-"Subtropics"
# 
# n<-which(metdata$Lat > 38)
# metdata$region[n]<-"North"
# 
# metdata$region<-as.factor(metdata$region)
# metdata<-subset(metdata, select=-WBAN)
# 
# ## load Hadley ISH raw data (only includes 1 zip file so far... add others!)
# load("/Users/elliotcohen/Dropbox/data/Electricity/CEA/Data/stnData.rsav")
# 
# ## remove erroneous values
# temps<-subset(stnData, select=-precip)
# temps<-subset(temps, temp > -100) # drop values below -100 deg. C
# temps<-subset(temps, temp < 200) # drop values above 200 deg. C (boiling)
# check(temps)
# temps$USAFID<-as.factor(temps$.id)
# 
# ## match metadat with observed data by USAFID
# df<-merge(temps, metdata, by="USAFID")
# 
# # ## match UN City data with Hadley ISD metadata...
# # ## use fuzzy logic...
# # get<-cities$City
# # grab<-metdata[which(metdata$City %in% get), ]
# 
# ## density plots
# sm.density.compare(df$temp, df$region, xlab="Temperature")
# title(main="Temperature Distribution by Region")
# colfill<-c(2:(2+length(levels(df$region)))) 
# legend("topleft", levels(df$region), fill=colfill)
```

Rationale
-----------------
Mid- to long-range global energy demand forecasts are typically reported as annual totals and provide little insight to the temporal distribution throughout the year. To address this shortcoming, we focus on the diurnal and seasonal distribution of energy demand and supply, which drive system cost but recieve relatively little attention in global energy outlooks. This requires more and better data than is typically available to researchers. 

A Global Survey
-------------------
This study presents an empirical analysis of diurnal-to-seasonal energy use patterns for XX global cities...

### Methods and Materials

Literature Review
-------------------

---
title: 'Global Trends In Urban Energy Use: The Tropical Shift'
author: "Elliot Cohen, Vijay Modi, Henri Torbey, Yu Tian and Michael Piccirilli"
date: "October 9 2014"
output:
  pdf_document: default
  html_document:
    theme: spacelab
---
